Pre working group data exploration: Mainly wrt coral cover

### 2024-07-10 julianna_data_exploration.Rmd
Broke beta diversity up by island side as well as habitat + year

```{r make long}
accum.HabISYrlong <- accumcomp.long(accum.HabISYr, ci=NA, 
                              label.freq=2000) %>% # lebel.freq adjusts the labelit variabel to show how often a symbol will be plotted on the SAC
  # get metadata back
  rowwise() %>% 
  mutate(Year = (str_split(Grouping, pattern = "_")[[1]])[length(str_split(Grouping, pattern = "_")[[1]])]) %>% 
  mutate(Habitat = case_when(
    str_split(Grouping, pattern = "_")[[1]][[3]] == "17" ~ "outer_17",
    str_split(Grouping, pattern = "_")[[1]][[3]] == "10" ~ "outer_10",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "fringing" ~ "fringing",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "backreef" ~ "backreef",
    TRUE ~ "Error"
    )) %>% 
  mutate(island_side = str_split(Grouping, pattern = "_")[[1]][[1]]) %>% 
  mutate(Habitat = factor(Habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) 
```

@@

```{r plot}
accum.HabISYrlong %>% 
  ggplot(aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) +
  scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) + # this moves it so zero is right at the origin
  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) + # this makes it more like a box
  geom_line(aes(colour=Year), linewidth=1) + # adds line
  #geom_point(data=subset(accum.Hablong, labelit==TRUE), 
  #             aes(colour=Grouping, shape=Grouping), size=.5) +
  #geom_ribbon(aes(colour=Year), alpha=0.2, show.legend=FALSE) + 
  BioR.theme +
  scale_colour_viridis_d(direction = -1) +
  facet_grid(island_side~Habitat) +
  labs(x = "Quadrats", y = "Richness")
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "fringing") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0fring
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "backreef") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0back
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "outer_10") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0outer10
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "outer_17") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0outer17
```

@@

## Plot together

How??

```{r}
```

@@

>>>>

```{r make long}
accum.HabISYrlong <- accumcomp.long(accum.HabISYr, ci=NA, 
                              label.freq=2000) %>% # lebel.freq adjusts the labelit variabel to show how often a symbol will be plotted on the SAC
  # get metadata back
  rowwise() %>% 
  mutate(Year = (str_split(Grouping, pattern = "_")[[1]])[length(str_split(Grouping, pattern = "_")[[1]])]) %>% 
  mutate(Habitat = case_when(
    str_split(Grouping, pattern = "_")[[1]][[3]] == "17" ~ "outer_17",
    str_split(Grouping, pattern = "_")[[1]][[3]] == "10" ~ "outer_10",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "fringing" ~ "fringing",
    str_split(Grouping, pattern = "_")[[1]][[2]] == "backreef" ~ "backreef",
    TRUE ~ "Error"
    )) %>% 
  mutate(island_side = str_split(Grouping, pattern = "_")[[1]][[1]]) %>% 
  mutate(Habitat = factor(Habitat, levels = c("fringing", "backreef", "outer_10", "outer_17")),
         island_side = factor(island_side, levels = c("north", "east", "west"))) 
```

@@

```{r plot}
accum.HabISYrlong %>% 
  ggplot(aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) +
  scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) + # this moves it so zero is right at the origin
  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) + # this makes it more like a box
  geom_line(aes(colour=Year), linewidth=1) + # adds line
  #geom_point(data=subset(accum.Hablong, labelit==TRUE), 
  #             aes(colour=Grouping, shape=Grouping), size=.5) +
  #geom_ribbon(aes(colour=Year), alpha=0.2, show.legend=FALSE) + 
  BioR.theme +
  scale_colour_viridis_d(direction = -1) +
  facet_grid(island_side ~ Habitat) +
  labs(x = "Quadrats", y = "Richness")
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year)) -> sppMeta0

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "fringing") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0fring
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "backreef") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0back
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "outer_10") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer10
```

@@

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
   filter(habitat == "outer_17") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer17
```

@@

## Join together

Attach back to metadata

```{r join distances with metadata}
sppMeta0fring %>% 
  mutate(distance_centriod = yrBetaDisperFring$distances) -> fringeDist

sppMeta0back %>% 
  mutate(distance_centriod = yrBetaDisperBack$distances) -> backDist

sppMeta0outer10 %>% 
  mutate(distance_centriod = yrBetaDisperOuter10$distances) -> outer10Dist

sppMeta0outer17 %>% 
  mutate(distance_centriod = yrBetaDisperOuter17$distances) -> outer17Dist
```

```{r join together}
# make sure this looks good
colnames(fringeDist) == colnames(backDist) 
colnames(backDist) == colnames(outer10Dist) 
colnames(outer10Dist) == colnames(outer17Dist)

# join together
betaDHabYr <- rbind(fringeDist, backDist, outer10Dist, outer17Dist)
```

Confirm these produce the same plots

```{r sanity check}
betaDHabYr %>% 
  ggplot(aes(x = year, y = distance_centriod, group = year)) +
  geom_boxplot() +
  facet_wrap(~habitat)
```

Add coral data

```{r}
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  # inner so that we don't add back in quadrats with 0% algae
  inner_join(betaDHabYr) -> betaDHabYrCoral
```

## Plot with coral

```{r scatter plot}
betaDHabYrCoral %>% 
  ggplot(aes(x = coral, y = distance_centriod)) +
  geom_point() +
  facet_grid(habitat ~ year)
```

```{r boxplot}
# get mean coral cover
betaDHabYrCoral %>% 
  group_by(year, habitat) %>% 
  summarize(Mean_coral = mean(coral)) %>% 
  full_join(betaDHabYrCoral) %>% 
  ggplot(aes(x = year, y = distance_centriod, group = year)) +
  geom_boxplot() +
  # Custom the Y scales:
  scale_y_continuous(
    # Features of the first axis
    name = "Algal beta diversity",
    # Add a second axis and specify its features
    sec.axis = sec_axis( trans=~.*100, name="Avg coral cover")
  ) +
  geom_line(aes(group = 1, x = year, y = Mean_coral/100), size = 2, color = "red") +
  facet_wrap(~habitat) 
```

```{r}
betaDHabYrCoral %>% 
  group_by(year, habitat) %>% 
  summarize(Mean_coral = mean(coral)) %>% 
  full_join(betaDHabYrCoral) %>% 
  ggplot(aes(group = 1, x = year, y = Mean_coral, group = year)) +
  geom_line()  +
  facet_wrap(~habitat) 
```

@@

## Dispersion by island side-year-habitat

Calculate beta dispersion (do it in groups for the sake of the computer)

```{r calculate beta dispersion by habitat}
yrISBetaDisperFring <- vegan::betadisper(brayMatFring, sppMeta0fring$island_side_year) # error: some squared distances are negative and changed to zero
# save(yrISBetaDisperFring, file = here("data/fringingyrISdispersion.Rdata"))

yrISBetaDisperBack <- vegan::betadisper(brayMatBack, sppMeta0back$island_side_year)  # error: some squared distances are negative and changed to zero
# save(yrISBetaDisperBack, file = here("data/backyrISdispersion.Rdata"))

yrISBetaDisperOuter10 <- vegan::betadisper(brayMatOuter10, sppMeta0outer10$island_side_year) # error: some squared distances are negative and changed to zero
# save(yrISBetaDisperOuter10, file = here("data/outer10yrISdispersion.Rdata"))

yrISBetaDisperOuter17 <- vegan::betadisper(brayMatOuter17, sppMeta0outer17$island_side_year) # error: some squared distances are negative and changed to zero
# save(yrISBetaDisperOuter17, file = here("data/outer17yrISdispersion.Rdata"))
```

```{r}
load(here("data/fringingyrISdispersion.Rdata"))
load(here("data/backyrISdispersion.Rdata"))
load(here("data/outer10yrISdispersion.Rdata"))
load(here("data/outer17yrISdispersion.Rdata"))
```

### Join together

```{r get all distances}
# get distances
sppMeta0fring %>% 
  mutate(distance_centriod = yrISBetaDisperFring$distances) -> fringeDistIS

sppMeta0back %>% 
  mutate(distance_centriod = yrISBetaDisperBack$distances) -> backDistIS

sppMeta0outer10 %>% 
  mutate(distance_centriod = yrISBetaDisperOuter10$distances) -> outer10DistIS

sppMeta0outer17 %>% 
  mutate(distance_centriod = yrISBetaDisperOuter17$distances) -> outer17DistIS
```

```{r join together}
# make sure this looks good
colnames(fringeDistIS) == colnames(backDistIS) 
colnames(backDistIS) == colnames(outer10DistIS) 
colnames(outer10DistIS) == colnames(outer17DistIS)

# join together
betaDHabYrIS <- rbind(fringeDistIS, backDistIS, outer10DistIS, outer17DistIS)

# add coral
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  # inner so that we don't add back in quadrats with 0% algae
  inner_join(betaDHabYrIS) -> betaDHabYrISCoral
```

### Plot

```{r without coral}
betaDHabYrISCoral %>% 
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
        habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) %>% 
  ggplot(aes(x = year, y = distance_centriod, group = year)) +
  geom_boxplot() +
  facet_grid(island_side~habitat)
```

With coral

```{r scatter plot}
betaDHabYrISCoral %>% 
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
        habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) %>% 
  ggplot(aes(x = coral, y = distance_centriod)) +
  geom_point() +
  facet_grid(habitat ~ year)
```

```{r boxplot}
# get mean coral cover
betaDHabYrISCoral %>% 
  group_by(year, habitat, island_side) %>% 
  summarize(Mean_coral = mean(coral)) %>% 
  full_join(betaDHabYrCoral) %>% 
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
        habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) %>% 
  ggplot(aes(x = year, y = distance_centriod, group = year)) +
  geom_boxplot() +
  # Custom the Y scales:
  scale_y_continuous(
    # Features of the first axis
    name = "Algal beta diversity",
    # Add a second axis and specify its features
    sec.axis = sec_axis( trans=~.*100, name="Avg coral cover")
  ) +
  geom_line(aes(group = 1, x = year, y = Mean_coral/100), size = 2, color = "red") +
  facet_grid(island_side~habitat) 
```
